---
title: "climateNyungwe"
author: "Murengera Aime Lucky Barahebuza"
date: "`r Sys.Date()`"
output: html_document
---

# How did temperature and precipitation change between 1970 - 2000 in Nyungwe National Park, Rwanda?

## Loading necessary packages and data
```{r setup, include=FALSE}
# library

library(geodata)
library(raster)
library(tidyverse)
library(ggplot2)

# loading temperature and precipitation data from WorldClim

temp_min <- worldclim_global(var = "tmin", res = 0.5, path = "01_data")
temp_max <- worldclim_global(var = "tmax", res = 0.5, path = "01_data")
temp_mean <- worldclim_global(var = "tavg", res = 0.5, path = "01_data")
prec <- worldclim_global(var = "prec", res = 0.5, path = "01_data")

# loading Nyungwe National Park shapefile
park_shp <- st_read("01_data/WDPA_WDOECM_Feb2025_Public_9148_shp_0/WDPA_WDOECM_Feb2025_Public_9148_shp-polygons.shp")

```

## plot the data first
```{r}
# Ploting Nyungwe National park

ggplot() +
  geom_sf(data = park_shp, fill = "darkgreen", color = "black") +
  theme_minimal() +
  labs(title = "Nyungwe National Park, Rwanda")

```

## How is the min temperature changing throught the year in Nyungwe?
```{r}
# crop raster files to the extent of Nyungwe National Park

nyungwe_crop <- raster::crop(temp_min, park_shp)
nyungwe_mask <- raster::mask(nyungwe_crop, park_shp)

# creating a dataframe to use

nyungwe_df <- as.data.frame(nyungwe_mask, xy = TRUE, na.rm = TRUE)
colnames(nyungwe_df)[3:14] <- month.abb
colnames(nyungwe_df)[1:2] <- c("longitude", "latitude")

# reshape the data to long format using tidyr

nyungwe_df <- pivot_longer(nyungwe_df, cols = Jan:Dec, names_to = "month", values_to = "temperature")

# converting month to an ordered factor

nyungwe_df$month <- factor(nyungwe_df$month, levels = month.abb)

# plot using ggplot

ggplot()+
  geom_tile(data = nyungwe_df, aes(x = longitude, y = latitude, fill = temperature))+
  geom_sf(data = park_shp, fill = NA, color = "black")+
  scale_fill_viridis_c( limits = c(8, 14))+
  facet_wrap(~month, ncol = 4)+
  theme_minimal()+
  labs(title = "Minimun Temperature in Nyungwe (1970 - 2000)", fill = "°C")

# plot linear trend

monthly_min_mean <- nyungwe_df |> group_by(month) |>
  summarise(mean_temperature = mean(temperature, na.rm = TRUE))

ggplot( monthly_min_mean, aes(x = month, y = mean_temperature, group = 1))+
  geom_line(color = "blue", linewidth = 1)+
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed")+
  theme_minimal()+
  labs(title = "Linear Monthly Temperature Trend", y = "Mean Temperature (°C)", x = "Month")

```

## How is the max temperature changing throught the year in Nyungwe?
```{r}
# crop raster files to the extent of Nyungwe National Park

nyungwe_maxt_crop <- raster::crop(temp_max, park_shp)
nyungwe_maxt_mask <- raster::mask(nyungwe_maxt_crop, park_shp)

# creating a dataframe to use

nyungwe_max_df <- as.data.frame(nyungwe_maxt_mask, xy = TRUE, na.rm = TRUE)
colnames(nyungwe_max_df)[3:14] <- month.abb
colnames(nyungwe_max_df)[1:2] <- c("longitude", "latitude")

# reshape the data to long format using tidyr

nyungwe_max_df <- pivot_longer(nyungwe_max_df, cols = Jan:Dec, names_to = "month", values_to = "temperature")

# converting month to an ordered factor

nyungwe_max_df$month <- factor(nyungwe_max_df$month, levels = month.abb)

# plot using ggplot

ggplot()+
  geom_tile(data = nyungwe_max_df, aes(x = longitude, y = latitude, fill = temperature))+
  geom_sf(data = park_shp, fill = NA, color = "black")+
  scale_fill_viridis_c( limits = c(16, 24))+
  facet_wrap(~month, ncol = 4)+
  theme_minimal()+
  labs(title = "Maximum Temperature in Nyungwe (1970 - 2000)", fill = "°C")
```

## How is the average temperature changing throught the year in Nyungwe?
```{r}
# crop raster files to the extent of Nyungwe National Park

nyungwe_avg_crop <- raster::crop(temp_mean, park_shp)
nyungwe_avg_mask <- raster::mask(nyungwe_avg_crop, park_shp)

# creating a dataframe to use

nyungwe_avg_df <- as.data.frame(nyungwe_avg_mask, xy = TRUE, na.rm = TRUE)
colnames(nyungwe_avg_df)[3:14] <- month.abb
colnames(nyungwe_avg_df)[1:2] <- c("longitude", "latitude")

# reshape the data to long format using tidyr

nyungwe_avg_df <- pivot_longer(nyungwe_avg_df, cols = Jan:Dec, names_to = "month", values_to = "temperature")

# converting month to an ordered factor

nyungwe_avg_df$month <- factor(nyungwe_avg_df$month, levels = month.abb)

# plot using ggplot

ggplot()+
  geom_tile(data = nyungwe_avg_df, aes(x = longitude, y = latitude, fill = temperature))+
  geom_sf(data = park_shp, fill = NA, color = "black")+
  scale_fill_viridis_c(limits = c(12, 20))+
  facet_wrap(~month, ncol = 4)+
  theme_minimal()+
  labs(title = "Average Temperature in Nyungwe (1970 - 2000)", fill = "°C")

```

# How is the average precipitation changing throught the year in Nyungwe?
```{r}
# crop raster files to the extent of Nyungwe National Park

nyungwe_prec_crop <- raster::crop(prec, park_shp)
nyungwe_prec_mask <- raster::mask(nyungwe_prec_crop, park_shp)

# creating a dataframe to use

nyungwe_prec_df <- as.data.frame(nyungwe_prec_mask, xy = TRUE, na.rm = TRUE)
colnames(nyungwe_prec_df)[3:14] <- month.abb
colnames(nyungwe_prec_df)[1:2] <- c("longitude", "latitude")

# reshape the data to long format using tidyr

nyungwe_prec_df <- pivot_longer(nyungwe_prec_df, cols = Jan:Dec, names_to = "month", values_to = "precipitation")

# converting month to an ordered factor

nyungwe_prec_df$month <- factor(nyungwe_prec_df$month, levels = month.abb)

# plot using ggplot

ggplot()+
  geom_tile(data = nyungwe_prec_df, aes(x = longitude, y = latitude, fill = precipitation))+
  geom_sf(data = park_shp, fill = NA, color = "black")+
  scale_fill_viridis_c(limits = c(12, 270))+
  facet_wrap(~month, ncol = 4)+
  theme_minimal()+
  labs(title = "Monthly Precipitation in Nyungwe (1970 - 2000)", fill = "mm")
```

# Yearly temperature trend between 1970-2000 in Nyungwe 
```{r}
library(raster)
library(ggplot2)

# Load monthly temperature data from WorldClim
tmean_stack <- stack("wc2.1_2.5m_tavg_*.tif")  # Replace with the correct path and files

# Calculate yearly average (assuming each 12 layers represent one year)
yearly_avg <- calc(tmean_stack, fun = function(x) rowMeans(matrix(x, nrow=12), na.rm=TRUE))

# Convert to a data frame for plotting
years <- 1970:2000  # Adjust based on your data
temp_df <- data.frame(Year = years, Temperature = yearly_avg[])

# Plot the trend
ggplot(temp_df, aes(x = Year, y = Temperature)) +
  geom_line(color = "blue") +
  geom_point() +
  labs(title = "Yearly Temperature Trend (WorldClim Data)", x = "Year", y = "Mean Temperature (°C)") +
  theme_minimal()

```
